A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-07-19 tidyindex.tex
plots for GGGI

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, breakable, sharp corners, enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

\hypertarget{an-example-on-sustainable-development-indicators}{%
\subsection{An example on sustainable development
indicators}\label{an-example-on-sustainable-development-indicators}}

In the following example, the Human Development Index (HDI) will first
be constructed using the pipeline syntax. The section will then
demonstrate how expressions or parameters in the pipeline can be swapped
to alternatives to study the index property.

Human Development Index (HDI) measures the development of countries
through more than just economic growth, but also people's life
expectancy and opportunity to receive education. These three identified
dimensions are measured using four indicators: life expectancy at birth
(health), expected years of schooling (education), mean years of
schooling (education), and Gross National Income (GNI) \emph{per capita}
(standard of living). The technical notes (United Nations Development
Programme 2021) have documented the procedures to calculate HDI and they
are summarised below:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  take log on GNI \emph{per capita},
\item
  rescale the four indicators into {[}0, 1{]} using mini-max,
\item
  aggregate the two education indicators using arithmetic mean, and
\item
  aggregate the three dimensions into the index using geometric mean.
\end{enumerate}

The values used in mini-max rescaling are summarised in
Table~\ref{tbl-rescale-params} and the justification of these numbers
can be found in the technical notes mentioned above.

\hypertarget{tbl-rescale-params}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2375}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.4500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1125}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1000}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1000}}@{}}
\caption{\label{tbl-rescale-params}Maximum and minimum values used to
rescale the four HDI indicators into {[}0, 1{]} range. The maximum of
GNI per capita is taken as the common log of 75,000, approximatly
4.875.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Dimension
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Indicator
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Minimum
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Maximum
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Dimension
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Indicator
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Minimum
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Maximum
\end{minipage} \\
\midrule()
\endhead
Health & Life expectancy at birth (years) & life\_exp & 20 & 85.000 \\
Education & Expected years of schooling (years) & exp\_sch & 0 &
18.000 \\
Education & Mean years of schooling (years) & avg\_sch & 0 & 15.000 \\
Standard of living & GNI per capita (2017 PPP\$) & gni\_pc & 2 &
4.875 \\
\bottomrule()
\end{longtable}

Among the four steps listed in the calculation, the first two are
variable transformation. The next two can be considered as the dimension
reduction step in the data pipeline given its intention to combine
multivariate data into univariate. With the index pipeline proposed in
the paper, HDI can be calculated as:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{dt }\OtherTok{\textless{}{-}}\NormalTok{ raw }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{init}\NormalTok{(}\AttributeTok{id =}\NormalTok{ country, }\AttributeTok{indicators =}\NormalTok{ life\_exp}\SpecialCharTok{:}\NormalTok{gni\_pc)}
\NormalTok{res }\OtherTok{\textless{}{-}}\NormalTok{ dt }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{var\_trans}\NormalTok{(}\AttributeTok{gni\_pc =} \FunctionTok{log}\NormalTok{(gni\_pc, }\AttributeTok{base =} \DecValTok{10}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{var\_trans}\NormalTok{(}\AttributeTok{.method =}\NormalTok{ rescale\_minmax, }\AttributeTok{.vars =}\NormalTok{ life\_exp}\SpecialCharTok{:}\NormalTok{gni\_pc,}
            \AttributeTok{min =} \FunctionTok{c}\NormalTok{(}\DecValTok{20}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{2}\NormalTok{), }\AttributeTok{max =} \FunctionTok{c}\NormalTok{(}\DecValTok{85}\NormalTok{, }\DecValTok{18}\NormalTok{, }\DecValTok{15}\NormalTok{, }\FunctionTok{log10}\NormalTok{(}\DecValTok{75000}\NormalTok{))) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{dim\_red}\NormalTok{(}\AttributeTok{sch =}\NormalTok{ (exp\_sch }\SpecialCharTok{+}\NormalTok{ avg\_sch) }\SpecialCharTok{/} \DecValTok{2}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{dim\_red}\NormalTok{(}\AttributeTok{index =}\NormalTok{ (life\_exp }\SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{*}\NormalTok{ gni\_pc)}\SpecialCharTok{\^{}}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\DecValTok{3}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

Index analysts can be interested in studying the property of the index
and the countries through the ``what-if'' questions. In the context of
HDI, this can be what if an alternative dimension reduction expression
is used to reduce the three dimensions into the HDI? Apart from the
geometrical mean, linear combination is a common method to reduce the
data dimension. With different weighting schemes, the combination can be
compared to study the variable importance of each dimension or to reveal
the underlying structures of countries. The following code tests five
expressions (arithmetic mean, principal component analysis weight, and
three weighting schemes emphasizing life expectancy, education, and GNI
per capita respectively) using the \texttt{swap\_exprs()} function. The
function first locates the operation to be tested
(\texttt{.var\ =\ index}), then accepts a list of expressions for
testing \texttt{.expr}, before requesting the raw data:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{(res2 }\OtherTok{\textless{}{-}}\NormalTok{ res }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{swap\_exprs}\NormalTok{(}
    \AttributeTok{.var =}\NormalTok{ index,}
    \AttributeTok{.expr =} \FunctionTok{list}\NormalTok{(}
      \AttributeTok{index1 =}\NormalTok{ (life\_exp }\SpecialCharTok{+}\NormalTok{ sch }\SpecialCharTok{+}\NormalTok{ gni\_pc)}\SpecialCharTok{/}\DecValTok{3}\NormalTok{,}
      \CommentTok{\# PCA recommended weight}
      \AttributeTok{index2 =} \FloatTok{0.569} \SpecialCharTok{*}\NormalTok{ life\_exp }\SpecialCharTok{+} \FloatTok{0.576} \SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{+} \FloatTok{0.586} \SpecialCharTok{*}\NormalTok{ gni\_pc, }
      \AttributeTok{index3 =} \FloatTok{0.8} \SpecialCharTok{*}\NormalTok{ life\_exp }\SpecialCharTok{+} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{+} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ gni\_pc,}
      \AttributeTok{index4 =} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ life\_exp }\SpecialCharTok{+} \FloatTok{0.8} \SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{+} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ gni\_pc,}
      \AttributeTok{index5 =} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ life\_exp }\SpecialCharTok{+} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{+} \FloatTok{0.8} \SpecialCharTok{*}\NormalTok{ gni\_pc),}
    \AttributeTok{.raw\_data =}\NormalTok{ dt))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
Index pipeline:

Summary: 
\end{verbatim}

\begin{verbatim}
var_trans: `log(gni_pc, base = 10)` -> gni_pc
\end{verbatim}

\begin{verbatim}
var_trans: `rescale_minmax(var = life_exp, min = 20, max = 85)` ->
life_exp
\end{verbatim}

\begin{verbatim}
var_trans: `rescale_minmax(var = exp_sch, min = 0, max = 18)` ->
exp_sch
\end{verbatim}

\begin{verbatim}
var_trans: `rescale_minmax(var = avg_sch, min = 0, max = 15)` ->
avg_sch
\end{verbatim}

\begin{verbatim}
var_trans: `rescale_minmax(var = gni_pc, min = 2, max = 4.875061)` ->
gni_pc
\end{verbatim}

\begin{verbatim}
dim_red: `(exp_sch + avg_sch)/2` -> sch
\end{verbatim}

\begin{verbatim}
dim_red: `(life_exp * sch * gni_pc)^(1/3)` -> index0
\end{verbatim}

\begin{verbatim}
dim_red: `(life_exp + sch + gni_pc)/3` -> index1
\end{verbatim}

\begin{verbatim}
dim_red: `0.569 * life_exp + 0.576 * sch + 0.586 * gni_pc` -> index2
\end{verbatim}

\begin{verbatim}
dim_red: `0.8 * life_exp + 0.1 * sch + 0.1 * gni_pc` -> index3
\end{verbatim}

\begin{verbatim}
dim_red: `0.1 * life_exp + 0.8 * sch + 0.1 * gni_pc` -> index4
\end{verbatim}

\begin{verbatim}
dim_red: `0.1 * life_exp + 0.1 * sch + 0.8 * gni_pc` -> index5
\end{verbatim}

\begin{verbatim}

Data: 
# A tibble: 191 x 13
      id country   life_exp exp_sch avg_sch gni_pc   sch index0 index1
   <dbl> <chr>        <dbl>   <dbl>   <dbl>  <dbl> <dbl>  <dbl>  <dbl>
 1     1 Switzerl~    0.984   0.917   0.924  0.983 0.920  0.962  0.963
 2     2 Norway       0.973   1       0.867  0.978 0.933  0.961  0.961
 3     3 Iceland      0.964   1       0.918  0.955 0.959  0.959  0.959
 4     4 Hong Kon~    1       0.960   0.815  0.973 0.887  0.952  0.953
 5     5 Australia    0.993   1       0.848  0.936 0.924  0.951  0.951
 6     6 Denmark      0.944   1       0.864  0.967 0.932  0.948  0.948
 7     7 Sweden       0.969   1       0.841  0.952 0.920  0.947  0.947
 8     8 Ireland      0.954   1       0.772  1     0.886  0.945  0.947
 9     9 Germany      0.933   0.945   0.939  0.952 0.942  0.942  0.942
10    10 Netherla~    0.949   1       0.839  0.956 0.919  0.941  0.941
# i 181 more rows
# i 4 more variables: index2 <dbl>, index3 <dbl>, index4 <dbl>,
#   index5 <dbl>
\end{verbatim}

The resulting data includes five more columns corresponding to the five
alternative indexes. The country ranking can be further computed on the
data and plotted in a scatterplot matrix as in
Figure~\ref{fig-hdi-expr}. The plot shows almost similar ranks suggested
by the indexes calculated using geometric mean (\texttt{rank0}), the
arithmetic mean (\texttt{rank1}), and the PCA recommended weights
(\texttt{rank2}). The agreement of the three methods supports the use of
geometric mean as the dimension reduction method in the original index.
The scatterplot matrix also reveals the difference in ranking when the
three dimensions are given different weights in \texttt{rank3},
\texttt{rank4}, and \texttt{rank5}. The panel \texttt{rank4} vs
\texttt{rank5} shows a cluster of highlighted points with relatively
high ranks according to \texttt{rank5} but low in \texttt{rank4}. These
countries are printed in Table~\ref{tbl-hdi-expr} prints along with
their three dimensions, calculated indexes, and rankings. While having a
high life expectancy and GNI \emph{per capita}, these countries
(Andorra, Qatar, San Marino, Kuwait, and Brunei Darussalam) score
relatively low in education, which is weighted heavily in
\texttt{index4}
(\texttt{0.1\ *\ life\_exp\ +\ 0.1\ *\ sch\ +\ 0.8\ *\ gni\_pc}).

@@

{\centering \includegraphics{tidyindex_files/figure-pdf/fig-hdi-expr-1.pdf}

@@

\caption{\label{fig-hdi-expr}Comparing country ranks from six different
index constructions: geometric mean (\texttt{rank0}), arithmetic mean
(\texttt{rank1}), principal component analysis recommended weights
(\texttt{rank2}), and heavier weight on life expectancy, education, and
GNI per capita (\texttt{rank3} - \texttt{rank5}). Geometric mean,
arithmetic mean, and PCA produce similar rankings where countries are
aligned in a diagnoal line, while country rankings varies when compare
among \texttt{rank3} to \texttt{rank5}. Panel \texttt{rank5}
vs.~\texttt{rank4} highlights a cluster of countries that rank high by
\texttt{rank5} but low by \texttt{rank4}. The plot demonstrates the
necessity to test on various alternative expressions to verify the
robustness of the method and to capture distinctive characteristics of
different countries.}

@@

\hypertarget{tbl-hdi-expr}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2118}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1882}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1176}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1765}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0824}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0824}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0706}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0706}}@{}}
\caption{\label{tbl-hdi-expr}A selected number of countries with low
rankings when education is given a heavy weight (\texttt{index4}) while
ranks high when a heavier weight is given on GNI per capital
(\texttt{index5}).}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
country
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
life expectancy
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
education
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
GNI per capita
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
index3
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
index4
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
rank3
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
rank4
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
country
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
life expectancy
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
education
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
GNI per capita
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
index3
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
index4
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
rank3
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
rank4
\end{minipage} \\
\midrule()
\endhead
Andorra & 0.929 & 0.721 & 0.942 & 0.909 & 0.764 & 32 & 69 \\
Qatar & 0.912 & 0.684 & 1.000 & 0.898 & 0.739 & 34 & 84 \\
San Marino & 0.937 & 0.701 & 0.947 & 0.914 & 0.749 & 30 & 76 \\
Kuwait & 0.903 & 0.670 & 0.947 & 0.884 & 0.721 & 39 & 96 \\
Brunei Darussalam & 0.841 & 0.694 & 0.977 & 0.840 & 0.737 & 60 & 85 \\
\bottomrule()
\end{longtable}

@@

\leavevmode\vadjust pre{\hypertarget{ref-UNDP_2021_tech_notes}{}}%
United Nations Development Programme. 2021. {``Human Development Report
2021-2022: The Next Frontier: Human Development and the Anthropocene.''}
\url{https://hdr.undp.org/sites/default/files/2021-22_HDR/hdr2021-22_technical_notes.pdf}.

\leavevmode\vadjust pre{\hypertarget{ref-hdi}{}}%
---------. 2022. {``Human Development Report 2021-22.''} New York.
\url{http://report.hdr.undp.org}.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, enhanced, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

\hypertarget{global-gender-gap-index}{%
\subsection{Global Gender Gap Index}\label{global-gender-gap-index}}

\includegraphics{tidyindex_files/figure-pdf/gggi-verify-1.pdf}

@@

{\centering \includegraphics{tidyindex_files/figure-pdf/gggi-weights-1.pdf}

@@

@@

{\centering \includegraphics{tidyindex_files/figure-pdf/idx-tour-1.pdf}

\caption{Five frames selected from varying the linear weights of four
dimensions in Global Gender Gap Index. The weights vary slightly from
the official simple average weights (0.25, 0.25, 0.25, 0.25) to observe
how the index and ranking reponse. Full animation is available at
{[}vimeo link{]}.}

@@

\leavevmode\vadjust pre{\hypertarget{ref-hdi}{}}%
United Nations Development Programme. 2022. {``Human Development Report
2021-22.''} New York. \url{http://report.hdr.undp.org}.
